Matroids \& testing in families of dependent hypotheses

### 2013-08-21 presentation/presentation.tex
streamlined shaffer procedure

\begin{document}
\frametitle{Matroids}

\maketitle

@@

\section{Algebraically dependent hypotheses}

\begin{frame}
\frametitle{Some history:}
\framesubtitle{Testing hypotheses of no differences in pairwise
  comparisons of 3 treatments}

@@

\begin{block}{$H_1:\mu_1=\mu_2,\; H_2:\mu_2=\mu_3,\;H_3:\mu_1=\mu_3$}
  \begin{enumerate}
  \item Test $H_ {\{1,2,3\}} = H_1\cap H_2 \cap H_3 :
    \mu_1=\mu_2=\mu_3$ at level $\alpha$,
  \item test each $H_1,H_2,H_3$ at level $\alpha$,
  \end{enumerate}
  protects the FWER at level $\alpha$.
\end{block}
\pause
  \begin{itemize}
    \item  essentially closed test, since %; no need to test pairwise
                                %intersection hypotheses since 
        $H_{\{i,j\}} = H_{\{1,2,3\}}, i\neq j,$

    \item  Bonferroni: if one of the hypotheses is rejected at level $\alpha/3$ the others can be tested at level alpha.
    \item  Hierarchical: if a predetermined primary hypothesis is rejected at level $\alpha$ the other two can be tested at level $\alpha$.
    \item  $\Rightarrow$ improves respective Bonferroni-Holm or fixed sequence tests.
  \end{itemize}

@@

    \input{shaffer}
\end{minipage} &
  \begin{minipage}{.6\textwidth}
    \begin{block}{$m=4, n=6, \only<2->{j=2}$}
      % \begin{itemize}
      % \item "Static": $t_j$ = max number of true hypotheses if $j$ are
      %   false.
      % \item "Dynamic": $t^*_j $ = max. number of true hypotheses if
      %   $H_{i_1},...,H_{i_j}$ are false.
%      \item Example:  \only<3->{$ j = 2$ :}
        \begin{description}
        \item<3->[Static:]  $H_1$ and $H_2$ are false
            $\Rightarrow$ $t_2 = 3$ (at most $H_4, H_5 $ and $H_6$ can be
            simultaneously true)
        \item<7->[Dynamic:] $H_1$ and $H_6$ are false $\Rightarrow$
          $t^*_2 = 2 < t_2 =3$ (only $H_3$, $H_4$, or $H_2$, $H_5$ can be
            simultaneously true)
        \end{description}

@@

\begin{frame}
\frametitle{Dependent null hypotheses}

\begin{itemize}
\item $H_i$ \alert{depends} on $H_I$: $H_{I \cup i} = H_{I}$
\item induces a \alert{dependence relation} $i \alert{\sim_H} I$ on index sets $I \in \mathcal{P}(S)$
\item $I$ \alert{independent} iff for all $i
  \in I$ $i \nsim_H I\setminus i$ (\ie $H_{I} \subsetneq H_{I\setminus i}$)
\pause
\item For pairwise comparisons $I$ independent iff there are no cycles
  in the corresponding graph
\item For general contrast tests $I$ independent iff $\bs{C_I}$ has full
 column rank
\end{itemize} % ends low level
\end{frame}

@@

\item Matroids generalize the structures of graphs and vector spaces,
 \eg cycles, bases, independent sets
\item A matroid $M$ on $S$ endows $\mathcal{P}(S)$ with an algebraic
 structure
\item \alert{Cryptomorphism} \ie many ways to define a matroid:
 \begin{itemize}
 \item From a graph perspective: Circuit axioms.
 \item From a vector space perspective: Independence axioms.
 \end{itemize}
%\item There is a dual matroid $M^*$ which can be helpful for proofs and algorithms
\end{itemize} % ends low level
\end{frame}

@@

\begin{frame}
 \frametitle{Example: Multiple doses}

\begin{tabular}{ll}
 \begin{minipage}{.4\textwidth}
   \begin{tikzpicture}[overlay]
     \node[yshift=70bp,xshift=-40bp] at (current page.south west){\input{lsd4.tex}};
   \end{tikzpicture}
 \end{minipage} &
 \begin{minipage}{.6\textwidth}
   \begin{tikzpicture}[remember picture,overlay]
     \node[yshift=100bp,xshift=160bp] at (current page.north west){\input{lsd4t.tex}};
   \end{tikzpicture}
 \end{minipage}
\end{tabular}
\end{frame}

>>>>

\begin{document}
\maketitle

@@

\section{Introduction}

\begin{frame}
\frametitle{Algebraic dependence}
\framesubtitle{Testing hypotheses of no differences in pairwise
  comparisons of 3 treatments}

@@

\begin{block}{$H_1:\mu_1=\mu_2,\; H_2:\mu_2=\mu_3,\;H_3:\mu_1=\mu_3$}
  \begin{enumerate}
  \item Test each $H_1,H_2,H_3$ at level $\alpha/3$.
  \item If any is rejected test the remaining at level $\alpha$.
  \end{enumerate}
  protects the FWER at level $\alpha$.
\end{block}
\pause
  \begin{itemize}
    \item  essentially closed test, since %; 
      \begin{enumerate}
        \item Test $H_ {\{1,2,3\}} = H_1\cap H_2 \cap H_3 :
          \mu_1=\mu_2=\mu_3$ at level $\alpha$,
        \item No need to test pairwise intersection hypotheses, as
          considered as subsets of the parameter space          
          $H_{\{i,j\}} = H_{\{1,2,3\}}, i\neq j$
      \end{enumerate}

    \item  $\Rightarrow$ improves respective Bonferroni-Holm test.
  \end{itemize}

\end{frame}
\begin{frame}
\frametitle{Contextual relationships}
\framesubtitle{Testing hypotheses of no differences in pairwise
  comparisons of 3 treatments}
\begin{block}{$H_1:\mu_1=\mu_2,\; H_2:\mu_2=\mu_3,\;H_3:\mu_1=\mu_3$}
  \begin{enumerate}
  \item Test $H_1$ at level $\alpha$.
  \item If $H_1$ is rejected, test $H_2,H_3$ at level $\alpha$ simultaneously.
  \end{enumerate}
  protects the FWER at level $\alpha$.
\end{block}
\pause
  \begin{itemize}
    \item  Uses a weighted Bonferroni test to reject $H_{1,2,3}$.
    \item Requires choosing a pre-fixed ``primary'' hypotheses. This
      depends on the context.
    \item  $\Rightarrow$ improves respective fixed sequence tests (no
      need to specify hierarchy for $H_2$ and $H_3$).
  \end{itemize}

@@

    \input{shaffer}
\end{minipage} &
  \begin{minipage}{.6\textwidth}
    \begin{block}{$m=4, n=6, j=2$}
      % \begin{itemize}
      % \item "Static": $t_j$ = max number of true hypotheses if $j$ are
      %   false.
      % \item "Dynamic": $t^*_j $ = max. number of true hypotheses if
      %   $H_{i_1},...,H_{i_j}$ are false.
%      \item Example:  \only<3->{$ j = 2$ :}
        \begin{description}
        \item<2->[Static:]  $H_1$ and $H_2$ are false
            $\Rightarrow$ $t_2 = 3$ (at most $H_4, H_5 $ and $H_6$ can be
            simultaneously true)
        \item<5->[Dynamic:] $H_1$ and $H_6$ are false $\Rightarrow$
          $t^*_2 = 2 < t_2 =3$ (only $H_3$, $H_4$, or $H_2$, $H_5$ can be
            simultaneously true)
        \end{description}

@@

\begin{frame}
\frametitle{Algebraically dependent null hypotheses}

\begin{itemize}
\item $H_j$ \alert{depends} on $H_I$: $H_{I \cup j} = \bigcap_{i \in
    I} H_i \cap H_j = \bigcap_{i \in I} H_i = H_{I}$
\item induces a \alert{dependence relation} $j \alert{\sim_H} I$ on index sets $I \in \mathcal{P}(S)$
\item $I$ \alert{independent} iff for all $i
  \in I$ $i \nsim_H I\setminus i$ (\ie $H_{I} \subsetneq H_{I\setminus i}$)
\pause
\item For pairwise comparisons $I$ independent iff there are no cycles
  in the corresponding graph
\item For general contrast tests $I$ independent iff $\bs{C_I}$ has full
 column rank
\end{itemize} % ends low level
\end{frame}

@@

\item Matroids generalize the structures of graphs and vector spaces.
\item A matroid $M$ on $S$ endows $\mathcal{P}(S)$ with an algebraic
 structure, \eg cycles, bases, independent sets, rank
\item \alert{Cryptomorphism} \ie many ways to define a matroid:
 \begin{itemize}
 \item From a graph perspective: Circuit axioms.
 \item From a vector space perspective: Independence axioms.
 \end{itemize}
%\item There is a dual matroid $M^*$ which can be helpful for proofs and algorithms
\end{itemize} % ends low level
\end{frame}

@@

\begin{frame}
 \frametitle{Example: QVA}
\begin{tabular}{ll}
 \begin{minipage}{.4\textwidth}
   \only<1,3>{   \begin{tikzpicture}[overlay]
       \node[yshift=100bp,xshift=-40bp] at (current page.south west){\input{qvameans.tex}};
     \end{tikzpicture}
   }   
   \only<2>{\begin{tikzpicture}[overlay]
       \node[yshift=-6cm,xshift=-1.4cm] at (current page.north west){\input{qvaalg.tex}};
     \end{tikzpicture}
   }
\end{minipage} &
 \begin{minipage}{.6\textwidth}
   \begin{block}{QVA Trial}
     \begin{itemize}
     \item QVA combination treatmend of QAB and NVA
     \item Three treatment groups against placebo (Pl)
     \item Fourth active comparator Tio
     \item Some secondary endponts SE
     \item<2-> If $H_1$ is rejected
     \item<2-> Test $H_2$, $H_4$ and $H_3$, $H_5$ simultaneously at level $\alpha/2$
     \item<2-> If only one is rejected no $\alpha$ may be passed on!
     \end{itemize}
   \end{block}
 \end{minipage}
\end{tabular}
\end{frame}

### 2013-08-21 presentation/shaffer.tex
streamlined shaffer procedure

\begin{tikzpicture}[scale=.4]
\node (H1) at (025bp,-125bp)[draw,circle,fill=white] {$\mu_1$};
\node (H2) at (225bp,-125bp)[draw,circle,fill=white] {$\mu_2$};
\node (H3) at (225bp,-325bp)[draw,circle,fill=white] {$\mu_3$};
\node (H4) at (025bp,-325bp)[draw,circle,fill=white] {$\mu_4$};
\only<-2>{\draw [line width=1pt] (H1) to (125bp, -124bp) node[fill=blue!20] {$H_{1}$} to (H2);}
\only<-2>{\draw [line width=1pt] (H1) to (075bp, -175bp) node[fill=blue!20] {$H_{2}$} to (H3);}
\only<3>{\draw [line width=1pt] (H1) to (125bp, -124bp) node[fill=red!20] {$H_{1}$} to (H2);}
\only<3>{\draw [line width=1pt] (H1) to (075bp, -175bp) node[fill=red!20] {$H_{2}$} to (H3);}
\only<-5>{\draw [line width=1pt] (H4) to (025bp, -227bp) node[fill=blue!20] {$H_{3}$} to (H1);}
\only<-4>{\draw [line width=1pt] (H2) to (225bp, -224bp) node[fill=blue!20] {$H_{4}$} to (H3);}
\only<-4>{\draw [line width=1pt] (H4) to (174bp, -176bp) node[fill=blue!20] {$H_{5}$} to (H2);}
\only<-4>{\draw [line width=1pt] (H3) to (126bp, -324bp) node[fill=blue!20] {$H_{6}$} to (H4);}
\only<4,5>{\draw [line width=2pt] (H2) to (225bp, -224bp) node[fill=green!20] {$H_{4}$} to (H3);}
\only<4,5>{\draw [line width=2pt] (H4) to (174bp, -176bp) node[fill=green!20] {$H_{5}$} to (H2);}
\only<4,5>{\draw [line width=2pt] (H3) to (126bp, -324bp) node[fill=green!20] {$H_{6}$} to (H4);}

\only<6>{\draw [line width=1pt] (H1) to (125bp, -124bp) node[fill=blue!20] {$H_{1}$} to (H2);}
\only<7>{\draw [line width=1pt] (H1) to (125bp, -124bp) node[fill=red!20] {$H_{1}$} to (H2);}
\only<6>{\draw [line width=2pt] (H3) to (126bp, -324bp) node[fill=blue!20] {$H_{6}$} to (H4);}
\only<7>{\draw [line width=2pt] (H3) to (126bp, -324bp) node[fill=red!20] {$H_{6}$} to (H4);}

\only<6,7,8,9>{\draw [line width=1pt] (H1) to (075bp, -175bp) node[fill=blue!20] {$H_{2}$} to (H3);}
\only<6,7>{\draw [line width=1pt] (H4) to (025bp, -227bp) node[fill=blue!20] {$H_{3}$} to (H1);}
\only<6,7>{\draw [line width=1pt] (H2) to (225bp, -224bp) node[fill=blue!20] {$H_{4}$} to (H3);}
\only<6,7,8,9>{\draw [line width=1pt] (H4) to (174bp, -176bp) node[fill=blue!20] {$H_{5}$} to (H2);}
\only<8>{\draw [line width=2pt] (H4) to (025bp, -227bp) node[fill=green!20] {$H_{3}$} to (H1);}
\only<8>{\draw [line width=2pt] (H2) to (225bp, -224bp) node[fill=green!20] {$H_{4}$} to (H3);}
\only<9>{\draw [line width=2pt] (H4) to (025bp, -227bp) node[fill=blue!20] {$H_{3}$} to (H1);}
\only<9>{\draw [line width=2pt] (H2) to (225bp, -224bp) node[fill=blue!20] {$H_{4}$} to (H3);}

\only<10>{\draw [line width=1pt] (H1) to (075bp, -175bp) node[fill=blue!20] {$H_{2}$} to (H3);}
\only<10,11,12>{\draw [line width=1pt] (H4) to (025bp, -227bp) node[fill=blue!20] {$H_{3}$} to (H1);}
\only<10,11,12>{\draw [line width=1pt] (H2) to (225bp, -224bp) node[fill=blue!20] {$H_{4}$} to (H3);}
\only<10>{\draw [line width=1pt] (H4) to (174bp, -176bp) node[fill=blue!20] {$H_{5}$} to (H2);}
\only<11,12>{\draw [line width=2pt] (H1) to (075bp, -175bp) node[fill=green!20] {$H_{2}$} to (H3);}
\only<11,12>{\draw [line width=2pt] (H4) to (174bp, -176bp) node[fill=green!20] {$H_{5}$} to (H2);}

>>>>

\begin{tikzpicture}[scale=.4]
\node (H1) at (025bp,-125bp)[draw,circle,fill=white] {$\mu_1$};
\node (H2) at (225bp,-125bp)[draw,circle,fill=white] {$\mu_2$};
\node (H3) at (225bp,-325bp)[draw,circle,fill=white] {$\mu_3$};
\node (H4) at (025bp,-325bp)[draw,circle,fill=white] {$\mu_4$};

\only<1,4->{\draw [line width=1pt] (H1) to (125bp, -124bp) node[fill=blue!20] {$H_{1}$} to (H2);}
\only<2-3,5->{\draw [line width=1pt] (H1) to (125bp, -124bp) node[fill=red!20] {$H_{1}$} to (H2);}

\only<1,4->{\draw [line width=1pt] (H1) to (075bp, -175bp) node[fill=blue!20] {$H_{2}$} to (H3);}
\only<2-3>{\draw [line width=1pt] (H1) to (075bp, -175bp) node[fill=red!20] {$H_{2}$} to (H3);}
\only<7->{\draw [line width=2pt] (H1) to (075bp, -175bp) node[fill=green!20] {$H_{2}$} to (H3);}

\only<1->{\draw [line width=1pt] (H4) to (025bp, -227bp) node[fill=blue!20] {$H_{3}$} to (H1);}
\only<6>{\draw [line width=2pt] (H4) to (025bp, -227bp) node[fill=green!20] {$H_{3}$} to (H1);}

\only<1-2,4-7>{\draw [line width=1pt] (H2) to (225bp, -224bp) node[fill=blue!20] {$H_{4}$} to (H3);}
\only<3,6>{\draw [line width=2pt] (H2) to (225bp, -224bp)  node[fill=green!20] {$H_{4}$} to (H3);}
\only<8>{\draw [line width=2pt,dashed] (H2) to (225bp, -224bp) node[fill=green!20] {$H_{4}$} to (H3);}

\only<1-2,4->{\draw [line width=1pt] (H4) to (174bp, -176bp) node[fill=blue!20] {$H_{5}$} to (H2);}
\only<3,7->{\draw [line width=2pt] (H4) to (174bp, -176bp) node[fill=green!20] {$H_{5}$} to (H2);}

\only<1-2,4->{\draw [line width=1pt] (H3) to (126bp, -324bp) node[fill=blue!20] {$H_{6}$} to (H4);}
\only<3>{\draw [line width=2pt] (H3) to (126bp, -324bp) node[fill=green!20] {$H_{6}$} to (H4);}
\only<5->{\draw [line width=1pt] (H3) to (126bp, -324bp) node[fill=red!20] {$H_{6}$} to (H4);}

% \only<6>{\draw [line width=1pt] (H1) to (125bp, -124bp) node[fill=blue!20] {$H_{1}$} to (H2);}
% \only<7>{\draw [line width=1pt] (H1) to (125bp, -124bp) node[fill=red!20] {$H_{1}$} to (H2);}
% \only<6>{\draw [line width=2pt] (H3) to (126bp, -324bp) node[fill=blue!20] {$H_{6}$} to (H4);}

% \only<6,7,8,9>{\draw [line width=1pt] (H1) to (075bp, -175bp) node[fill=blue!20] {$H_{2}$} to (H3);}
% \only<6,7>{\draw [line width=1pt] (H4) to (025bp, -227bp) node[fill=blue!20] {$H_{3}$} to (H1);}

% \only<6,7,8,9>{\draw [line width=1pt] (H4) to (174bp, -176bp) node[fill=blue!20] {$H_{5}$} to (H2);}

% \only<8>{\draw [line width=2pt] (H2) to (225bp, -224bp) node[fill=green!20] {$H_{4}$} to (H3);}
% \only<9>{\draw [line width=2pt] (H4) to (025bp, -227bp) node[fill=blue!20] {$H_{3}$} to (H1);}
% \only<9>{\draw [line width=2pt] (H2) to (225bp, -224bp) node[fill=blue!20] {$H_{4}$} to (H3);}

% \only<10,11,12>{\draw [line width=1pt] (H4) to (025bp, -227bp) node[fill=blue!20] {$H_{3}$} to (H1);}

% \only<10>{\draw [line width=1pt] (H4) to (174bp, -176bp) node[fill=blue!20] {$H_{5}$} to (H2);}
% \only<11,12>{\draw [line width=2pt] (H1) to (075bp, -175bp) node[fill=green!20] {$H_{2}$} to (H3);}
% \only<11,12>{\draw [line width=2pt] (H4) to (174bp, -176bp) node[fill=green!20] {$H_{5}$} to (H2);}
